\section{Introduction}\label{sec:intro}

The Andrews--Curtis (AC) conjecture \cite{andrews1965ac} states that every \emph{balanced}
presentation of the trivial group --- a presentation $\langle x_1,\dots,x_n \mid
r_1,\dots,r_n\rangle$ with as many relators as generators that presents the one-element
group --- can be transformed into the trivial presentation $\langle x_1,\dots,x_n \mid
x_1,\dots,x_n\rangle$ by a finite sequence of elementary moves. Sixty years after it was
posed, it remains open, and its difficulty is not merely combinatorial folklore: a
balanced presentation that resisted trivialization would refute the conjecture, and the
same balanced presentations sit at the heart of low-dimensional topology, where they
encode handle decompositions of contractible $4$-manifolds. Potential AC counterexamples
arise directly from Akbulut--Kirby's smooth $4$-dimensional constructions
\cite{akbulut1985ac} and from candidate counterexamples to the Property $2R$ and
slice--ribbon conjectures \cite{gompf2010fibered}, and the conjecture is explicitly
entangled with the smooth $4$-dimensional Poincar\'e problem \cite{freedman2010poincare}.
Among all candidates, the Akbulut--Kirby presentation $\mathrm{AK}(3) = \langle x,y \mid
x^{3}=y^{4},\ xyx=yxy\rangle$ is the shortest and most-studied potential counterexample,
standing since 1985.

The \emph{stable} (or \emph{weak}) AC conjecture relaxes the problem by allowing two
further moves, stabilization by a trivial generator--relator pair (AC4) and its inverse
(AC5); a presentation reducible to the trivial one under AC1--AC5 is \emph{stably
AC-trivial}. The stable status of $\mathrm{AK}(3)$ has recently oscillated. The first
version of a 2024 case study \cite{shehper2024hard} announced that $\mathrm{AK}(3)$ is
stably AC-trivial, removing it as a stable-AC counterexample; a subsequent revision
located a misprint in the Wirtinger construction of \cite{myasnikov2002ac} on which that
argument depended and rescinded the claim. A parallel automated-deduction paper
\cite{lisitsa2025ak3}, written between those two versions, builds its stable-triviality
conclusion on the same, now-rescinded premise and has not been revised. The plain reading
of the current literature is therefore blunt: the stable AC-triviality of $\mathrm{AK}(3)$
is open again.

Search-based and learning-based methods have crossed hard barriers in this domain before
--- genetic algorithms trivialized $\mathrm{AK}(2)$ \cite{miasnikov1999genetic}, and a
recent study of the AC difficulty landscape \cite{fagan2026twohump} pushed reinforcement
learning past presentations that classical search cannot reach. This invites a concrete
question about stabilization as an escape hatch: does adding a generator $z$ defined by a
relator $z^{-1}w(x,y)$ --- for a cleverly chosen word $w$ in the original generators --- open
shortcuts (informally, ``wormholes'') through the enlarged AC graph that ordinary
2-generator search cannot find? If some $w$ trivialized $\mathrm{AK}(3)$ where the dumb
choices $z=r_i$ do not, stabilization would be a genuine algorithmic lever rather than
bookkeeping.

This paper is a systematic audit of that hope, and it returns a rigorously verified
negative paired with a structural map (Figure~\ref{fig:overview}). We first re-verify,
seven independent ways, that the stable status of $\mathrm{AK}(3)$ is open while its
plain-AC equivalence to the length-25 presentation P25 survives. We then run three
stabilization strategies of increasing sophistication --- all of which fail within the
searched budgets --- and a control showing that the per-relator length cap is not what stops
them. Finally, we characterize \emph{where} stabilized greedy search stalls: a length-13
floor whose AC-class has exactly two canonical representatives, bridged by a certified
move sequence. Throughout, every positive claim is emitted as a machine-checkable
certificate.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  font=\small,
  box/.style={draw, rounded corners, align=center, inner sep=5pt, fill=gray!12},
  attack/.style={draw, rounded corners, align=center, text width=33mm,
                 minimum height=11mm, inner sep=4pt, fill=gray!3},
  arr/.style={-{Latex[length=2.2mm]}, thick, gray!55},
]
\node[box, text width=30mm] (m)
  {\textbf{MMS02 misprint}\\$\to$ v2 rescission\\[2pt]$\mathrm{AK}(3)$ stable\\status re-opened\\(7/7 checks)};

\node[attack, right=17mm of m, yshift=15mm] (a1) {naive $z{=}w$ over 640 MS};
\node[attack, below=5mm of a1] (a2) {97-word bank on $\mathrm{AK}(3)$};
\node[attack, below=5mm of a2] (a3) {five lanes incl.\ Lemma-11 plateau elim.:\\180{,}645 quotients};

\node[draw, dashed, rounded corners, fit=(a1)(a2)(a3), inner sep=6pt,
      label={[font=\footnotesize]above:three stabilization attacks}] (grp) {};

\node[align=center, font=\footnotesize\bfseries, right=11mm of a2] (zero)
  {0 solved\\(within budget)};

\node[box, text width=24mm, right=8mm of zero] (cap)
  {\textbf{cap control}\\$L\le48$:\\not binding};

\node[box, text width=33mm, right=8mm of cap, fill=gray!22] (floor)
  {\textbf{length-13 floor}\\two canonical reps\\$F$ (71\%) / $\mathrm{AK}(3)$ (29\%)\\certified 21-move bridge};

\draw[arr] (m) -- (grp);
\draw[arr] (a1.east) -- (zero.north west);
\draw[arr] (a2.east) -- (zero.west);
\draw[arr] (a3.east) -- (zero.south west);
\draw[arr] (zero) -- (cap);
\draw[arr] (cap) -- (floor);
\end{tikzpicture}%
}
\caption{End-to-end audit pipeline: a greedy substitution solver, a stabilization move
that defines a fresh generator $z$ by a chosen word $w(x,y)$, a $97$-word
literature-grounded bank, and five attack lanes (meet-in-the-middle, direct stable-move
search, trivial-$z$ control, plateau elimination, and a zero-shot RL beam probe), with
every reported solution emitted as a machine-checkable certificate replayed by two
independent verifiers.}
\label{fig:overview}
\end{figure}

\paragraph{Contributions.}
\begin{itemize}
  \item \textbf{Re-verification of a re-opened question.} We computationally confirm, via
    7 of 7 agreeing independent checks (Smith normal form, Todd--Coxeter coset enumeration,
    and $S_3$-homomorphism counting among them), that $\mathrm{AK}(3)$'s stable status is
    open, that the plain-AC link $\mathrm{AK}(3)\sim\mathrm{P}25$ survives (53-move replay,
    plus validation of an independent Prover9 derivation), and we surface two data defects
    in a published trivialization artifact.
  \item \textbf{Systematic stabilization negatives at stated budgets.} Across a naive
    $z=w$ construction over 640 Miller--Schupp presentations, a 97-word literature-grounded
    bank on $\mathrm{AK}(3)$, and a five-lane campaign that mines 180{,}645 distinct
    2-generator quotients, 0 of 16{,}870 solve attempts succeed within the searched
    budgets (up to $2\times10^{6}$ nodes per search, beam width $\le 2048$).
  \item \textbf{A cap-not-binding control.} We show the $L=24$ per-relator length cap is
    not the binding constraint in the tested range: relaxing to $L\le 48$ (and total
    length $\le 60$) does not convert a single failure into a success.
  \item \textbf{A two-representative structural floor.} The length-13 floor of
    $\mathrm{AK}(3)$'s AC-class has exactly two canonical (signed-relabel) representatives
    --- $F=\langle x,y \mid y^{-2}xyx^{-2},\ y^{-3}x^{-2}yx\rangle$, a $2$-generator
    presentation AC-equivalent to $\mathrm{AK}(3)$ (712/1006 $=71\%$ of floor states), and
    $\mathrm{AK}(3)$'s own reduced form (294/1006 $=29\%$) --- joined by a certified 21-move
    AC path.
  \item \textbf{A certificate framework and reusable archive.} We release the acx-cert-v1
    machine-checkable certificate format with two independently authored verifiers
    (20{,}947 checks, 0 failures) and the full candidate-quotient archive, so every claim
    above is independently re-runnable.
\end{itemize}

Code, data, and all certificates will be released on acceptance.
