\section{Conclusion and Next Steps}\label{sec:conclusion}

AK(3)'s stable AC-triviality remains open: the premise that had appeared to settle it was
independently re-verified as rescinded in prior work \cite{shehper2024hard}. Within the
searched budgets ($\leq 2\times10^6$ nodes per run, beam width $\leq 2048$), stabilizing AK(3)
in every tested form adds nothing over plain 2-generator greedy substitution --- naive $z=w$
arms solve strict subsets of what the unstabilized baseline already solves, no word in the
97-word bank beats a target's own relators as a stabilizer, and across five
differently-shaped search lanes, 0 of 16,870 attempts solved. The relator-length cap does
not bind in the tested range ($L\leq48$, solve cap 60): raising it changes nothing. And the
entire accessible region funnels onto a length-13 floor with exactly two canonical
(signed-relabel) representatives of that floor --- $F$, a 2-generator presentation
AC-equivalent to AK(3) (71\% of the census), and AK(3)'s own reduced form (29\%) --- joined by
a certified 21-move AC path (\ref{fig:two_floors}). That two-floor result, not a solve, is
this paper's certified structural contribution: a new fact about AK(3)'s neighborhood,
established independently of whether AK(3) itself ever trivializes.

The negative and the floor structure point at the same mechanism. Greedy substitution,
ordered by shrinking total length, cannot assemble a Lemma-11 change-of-variables one
length-decreasing move at a time, because the cancellation the lemma provides is inherently
a batch effect, not a gradual descent --- the useful move is an atomic destabilization
supermove, not incremental substitution. The literature's own 53-move trivialization path
confirms this directly: it peaks at total length 25 before falling to trivial, climbing
before it descends. Read this way, the two-hump landscape reported for the unstable problem
\cite{fagan2026twohump} reappears here one level up --- a second hump, for the stable problem,
separating the length-13 floor from the trivial presentation, that only a search capable of
climbing (not just shrinking) has any chance of crossing.

\paragraph{Next steps.}
\begin{enumerate}
  \item Make Lemma-11 elimination itself a first-class, length-increasing search move,
    gated against the certified 21-move chain; the StableSolver machinery already exists
    (\ref{sec:methods}) and needs scaling, not redesign.
  \item Attack $F$ directly. It is the dominant floor representative (71\% of the census)
    and, unlike AK(3), was never itself used as a search seed before this work; inverting
    the certified $F\to$AK(3) bridge gives a ready-made route to search outward from it.
  \item Train or fine-tune the RL policy on floor quotients and stabilized states directly,
    closing the zero-shot out-of-distribution gap noted in \S5.
  \item Escalate the archived 180,645-quotient candidate pool at 10--40$\times$ the tested
    budgets; its deduplicated, resumable construction (\ref{app:repro}) makes this a scaling
    exercise, not a rerun from scratch.
  \item Test non-length-ordered or globally-guided search; the Appendix-F path's length-25
    peak is a concrete target phenomenon that a length-ordered search cannot produce by
    construction.
  \item Build a per-presentation word-bank generator that generalizes the
    target-derived-relator construction used for the hard-target controls, so word choice no
    longer depends on AK($n$)-specific theory.
  \item Export certificates for the remaining 137 catalog-leaf classifications and extend
    the independent verifier to cover the harvest pipeline end to end, not only the
    certificates already exported.
\end{enumerate}

Negative results of this breadth, reproducible and certificate-grade, carve the search
space for anyone who takes up this problem next. The floor census, the certificate chains,
and the deduplicated candidate archive are released, on acceptance, as a foundation for
exactly that.
