\section{Reproducibility and Artifact Details}\label{app:repro}

\subsection{Data and artifacts}

\begin{itemize}
  \item \textbf{MS(1190) dataset.} The $640$-presentation known-solvable prefix used
    throughout \ref{sec:methods}, drawn from the Miller--Schupp benchmark
    \cite{miller1999ms}.
  \item \textbf{Consolidated candidate/trial archives.} JSONL.gz streams with one record
    per (presentation, arm/word/candidate, budget) triple and full per-record provenance ---
    source form, generator and relator words, node count, floor reached, solved flag.
  \item \textbf{Certificates.} JSON, schema of \ref{app:certs}.
\end{itemize}

All of the above, together with the two acx-cert-v1 verifiers, are released on acceptance.
The source MS(1190)/AC-benchmark datasets this campaign builds on are released under a data
license (CC BY 4.0) by \cite{fagan2026twohump}; the campaign's own derived artifacts --- word
bank, quotient archive, certificates --- will be released under the same terms.

\subsection{Per-campaign recipe}

Six campaigns, each an independent, resumable, append-only JSONL stream keyed by (dataset,
idx-or-form, word-or-arm, budget):

\begin{enumerate}
  \item \textbf{Baseline reproduction} (cap-semantics check, \ref{sec:methods}). Entry:
    \texttt{run\_greedy.py}. Parameters: \texttt{--cap \{sum,per\_relator\} --budget
    \{1e5,1e6\} --workers N}. Output: one stream per (cap, budget) tier, one record per
    presentation index, distinguishing the per-relator cap ($L=24$) from the original sum
    cap.
  \item \textbf{Naive $z=w$ arms} (Table~\ref{tab:arms}). Entry:
    \texttt{run\_greedy\_sweep.py}. Parameters: arm $\in\{$baseline$,r_1,r_2,x,y\}$, budget
    $5\times10^5$, two-phase light/heavy worker scheduling so cheap instances finish first
    and memory-heavy ones run at low parallelism. Output: one stream per arm, records keyed
    by (dataset, idx, arm, budget) so arms merge directly for the subset/union comparison
    of \ref{sec:results}.
  \item \textbf{AK(3) word bank} (Table~\ref{tab:wordbank}). Entry:
    \texttt{run\_ak3\_wormhole.py}. Parameters: form $\in\{$textbook, rep$\}$, the 97-word
    bank, a two-tier budget ($10^5$-node screen, escalate every unsolved word to $10^6$),
    per-relator cap $L=24$. Output: one stream per (form, budget), named by the pattern
    \texttt{ak3\_<form>\_<budget>.jsonl}, keyed by word name.
  \item \textbf{Hard-target controls} (Figure~\ref{fig:hard_ties}). Entry:
    \texttt{run\_hard\_wormhole.py}. Parameters: MS(1190) indices $\{625,610\}$,
    target-specific word banks (the \texttt{relhalf} family re-derived from each target's
    own relators), $10^5$-node screen. Output: one stream per target index, keyed by word
    name.
  \item \textbf{Five-lane campaign} (Table~\ref{tab:lanes}). Entries:
    \texttt{plateau\_elim.py} (Lane D's harvest/merge/solve phases), with
    \texttt{harvest\_fast.py} as its numba-accelerated, byte-identical hot path;
    \texttt{stable\_solver.py} (Lane B); \texttt{floor\_census.py} (the two-floor census);
    \texttt{resolve\_hi\_l.py} (cap-gap re-solve of the shortest candidates at a raised
    per-relator cap); \texttt{collect\_results.py}, which consolidates every box's streams
    into one archive plus a manifest. Parameters vary by phase: harvest budgets up to
    $\sim\!5\times10^5$ per combination, solve budgets $2.5\times10^4$--$2\times10^5$,
    \texttt{resolve\_hi\_l.py} at cap $L=40$. Output: per-combination candidate streams, a
    global merged-candidate stream, a solve-attempt stream, and certificate JSON files ---
    keyed by (form, word) for harvest and by candidate canonical key for merge/solve.
  \item \textbf{Cap control} (\ref{app:cap}). Entries: \texttt{run\_captest.py}
    (positive-control gate, the search-cap arm, and the harvest-cap arm) and
    \texttt{solve\_stratified.py} (the stratified-solve arm). Parameters: gate at
    $5\times10^4$ nodes; search arm at $\max\_len\in\{24,48\}$, $10^5$ nodes; harvest arm at
    per-relator cap $48$, total-length cap $40$, $6\times10^4$ nodes; stratified solve at
    $100$ candidates per length bucket $25$--$40$ plus $200$ short controls,
    $\max\_len=60$, budget $8\times10^4$ ($4\times10^4$ for the longest-tail bucket).
    Output: isolated JSONL streams, kept entirely separate from the main campaign's
    directories.
\end{enumerate}

Every stream is append-only, one JSON record per line, flushed (and fsynced on
cloud/networked mounts) after each write. A re-run of a finished sweep re-reads existing
ids and exits as a no-op; a killed run resumes from the last complete line, discarding at
most one in-flight record.

\subsection{Determinism and verification}

The per-node hot paths --- neighbor generation, canonicalization, Lemma-11 elimination --- are
numba-jitted. Every jitted path is held byte-identical to a pure-Python reference
implementation via differential tests on real visited sets, with $0$ mismatches across the
tested combinations. Every claimed solve is replayed through an independent path verifier
before it is counted (Table~\ref{tab:certs}, \ref{app:certs}), and every printed number in
the paper is re-derived by an independent recomputation script run directly against the raw
JSONL/gz streams, not against any intermediate digest.

\subsection{Compute}

Development and calibration ran on a single consumer laptop ($16$ GB RAM); the five-lane
escalation additionally used cloud CPU boxes (up to $8$ vCPU, up to $50$ GB RAM); the RL
beam probe (Lane E) ran as CPU-only inference throughout. Per-attempt node budgets ranged
from $2.5\times10^4$ to $10^6$; a greedy run's visited set carries roughly $40\times$ its
node budget in entries, which is the binding memory constraint at the higher budgets. Total
wall-clock across the full campaign was approximately one week.
